---
title: "PSY6422 Data Management and Visualisation"
output:
  html_document:
    df_print: paged
    fig_width: 7
    fig_height: 6.37
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## The Growth of Woodland Area within the United Kingdom (1998-2023)

### Data Source

This project is focussed on visualising the increase in woodland within the **United Kingdom**, as categorised by each *country*, from *1998* to *2023*. The raw data for this project was retrieved from ['Forest Research'](https://www.forestresearch.gov.uk/tools-and-resources/statistics/data-downloads/), a branch of the Forestry Commission (FC) which collects data by totaling extractions made by multiple forest services operating in the United Kingdom. It principally gathers information on tree-related statistics, including but not limited to gathering longitudinal provisional data on the estimated amount of woodland area in the UK.

### The Research Question

The main rationale for exploring this data was to elucidate whether there has been a growth in woodland area within the United Kingdom between 1998 and 2023, using existing data on the amount of hectares of woodland within each country. In tandem, calculating the *percentage* increase per country would aid in contextualising whether countries are developing at similar or dissimilar rates, in spite of notably different amounts of actual woodland.

### Data Preparation

The “Woodland area, UK, 1998 to 2023” ODS file contained 5 sheets of actual data, though was also supplemented with cover, content and notes pages. In tandem, the pages containing data were annotated with further notes clarifying the origin of the data and different commissions involved in gathering it. 

	
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Loading of necessary libraries

library(tidyverse)
library(ggplot2)
library(readODS)
library(here)
library(rlang)
library(dplyr)
library(cowplot)
library(gridExtra)

# Extraction of raw data

sheets<-c("England","Wales","Scotland","Northern Ireland")
countries_list<-list()
for (i in seq_along(sheets)){
  countries<-sheets[i]
  pathway<-paste0(here("raw_data", "area-timeseries-15jun23.ods"))
  countries_list[[countries]]<-read_ods(pathway, sheet = i+3)
} # extracts all of the raw data and collates it into a data-frame

raw_extracted_data<-data.frame(countries_list)
rm(list=setdiff(ls(), "raw_extracted_data")) # removes unnecessary variables to clean environment

raw_extracted_data
```



Before continuing, the unnecessary pages and lines of text needed to be removed in order to make the data more amenable to analysis.

```{r, echo=TRUE}
# Processing and cleaning of data

processed_extracted_data<-raw_extracted_data[-c(1:3),] # removes unnecessary text
names(processed_extracted_data)<-as.matrix(processed_extracted_data[1,]) # labels each column by their original titles
processed_extracted_data<-processed_extracted_data[-1,] # removes text names from the data

processed_extracted_data
```





From all of the data, only the *Year*, and the *Woodland Area* (in thousand hectares) of each country needed to be extracted. All of the data on the amount of woodland area also needed to be used to calculate the percentage increase from 1998 to 2023 for each country.

```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Extracts the Years and Amount of Woodland Area for each country

year_ending_March_31st<-as.numeric(processed_extracted_data$`Year ending 31 March`) # The Year
woodland_area<-select(processed_extracted_data, ends_with(" total (thousand ha)")) # Amount of Woodland Area
country_names<-c("England", "Wales", "Scotland", "Northern Ireland") # Country Names

for (i in country_names){
  assign(i, as.numeric(woodland_area[[paste0(i, " total (thousand ha)")]])) 
} # converts data to numeric for visualisation

# Tabulates change in woodland area per country from 1998 to 2023

woodland_area_by_country<-data.frame(England,Wales,Scotland,`Northern Ireland`)
woodland_area_by_country<-woodland_area_by_country%>%
  rename(Northern_Ireland=`Northern.Ireland`)
table(is.na(woodland_area_by_country)) # no missing data!
country_names<-c("England", "Wales", "Scotland", "Northern_Ireland")

# Calculates percentage increase in woodland area per country from 1998 to 2023

percentage_results<-data.frame(country=character(),percentage_increase=numeric(),stringsAsFactors=FALSE)
for(country in country_names){
  min_v<-min(woodland_area_by_country[[country]])
  max_v<-max(woodland_area_by_country[[country]])
  percentage_increase<-(((max_v-min_v)/min_v)*100)
  percentage_results<-rbind(percentage_results, data.frame(country=country,percentage_increase=percentage_increase))
} # calculates the percentage increase from 1998 to 2023 for each country
percentage_results[1:4,2]<-round(percentage_results[1:4,2],2) # rounds the data to 2 decimal places

# Data check 

woodland_area_by_country
percentage_results
```

### Data Visualisation

Finally, a visualisation of the growth of woodland area within the countries of the United Kingdom from 1998 to 2023 can be created, which displays this development both in terms of numeric and proportional increase.

```{r, message=FALSE, echo=TRUE}
# Removal of all unnecessary variables, and creation of data-frame amenable to the upcoming visualisation

rm(list=setdiff(ls(),c("woodland_area_by_country","country_names","year_ending_March_31st", "plot","percentage_results")))
woodland_growth_over_time<-data.frame(
  country=c(rep("England",26),rep("Wales",26),rep("Scotland",26),rep("Northern Ireland",26)),
  woodland=c(woodland_area_by_country$England,woodland_area_by_country$Wales,
             woodland_area_by_country$Scotland,woodland_area_by_country$Northern_Ireland),
  year=c(year_ending_March_31st)
)

# Mapping & figure path

mapping<-aes(x=year,y=woodland,colour=country) # creates the mapping for the visualisation
fig_path<-here("figs") # creates the necessary path for saving the figure

# Visualisation of the growth of woodland area within the United Kingdom, from 1998 to 2023

FinalPlot<-woodland_growth_over_time %>%
  ggplot(mapping=mapping)+
  geom_smooth(method="gam")+
  geom_point()+
  labs(x="Year commencing from March 31st",
       y="Woodland area (in thousand hectares)",
       colour="Country",
       title="The Growth of Woodland Area within the United Kingdom",
       subtitle="As annotated with percentage increase from 1998-2023",
       caption="Retrieved from: Forest Research, 2023")+
  annotate("label",x=2010,y=1210,label=paste(percentage_results[1,2],"%"),colour="#EE0000",size=3,fontface="bold")+
  annotate("label",x=2010,y=375,label=paste(percentage_results[2,2],"%"),colour="#00CD00",size=3,fontface="bold")+
  annotate("label",x=2010,y=1450,label=paste(percentage_results[3,2],"%"),colour="#0000CD",size=3,fontface="bold")+
  annotate("label",x=2010,y=175,label=paste(percentage_results[4,2],"%"),colour="#FFA500",size=3,fontface="bold")+
  scale_colour_manual(values=c(England="#EE0000",Wales="#00CD00",
                               Scotland="#0000CD",`Northern Ireland`="#FFA500"))+
  scale_x_continuous(breaks=seq(1998,2023,5))+
  scale_y_continuous(breaks=seq(0,1500,150))+
  theme(panel.border=element_rect(colour="#8B7355",fill=NA,linewidth=2),
        panel.grid.minor=element_line(colour="#CAFF70"),
        panel.grid.major=element_line(colour="#CAFF70"),
        panel.background=element_rect(fill="white"))

FinalPlot
```

### Saving the Plot

```{r, message=FALSE, echo=TRUE}
filename<-paste("The Growth of Woodland Area in the UK from 1998 to 2023.png",sep="")
ggsave(file.path(fig_path,filename),plot=FinalPlot,width=7,height=6.37)
```

### Conclusion

From this plot, there is visual clarity that the growth of woodland area within countries should be interpreted both in the context of actual hectares of woodland area developed, in tandem with the proportional increase relative to the original amount of woodland within the country - this is because taking either of these measures outside of the context of the other could be misleading. For example, interpreting that Wales has made superior efforts to increase Woodland area compared to Northern Ireland, in the context that Wales visually has a higher *amount* of woodland, neglects how Northern Ireland has theoretically made grander efforts than Wales to expand on the notion that they have a higher *percentage increase* in woodland. Regardless of these differences, the data can simply be interpreted on the promising notion that all countries within the United Kingdom have experienced an expansion of woodland between 1998 and 2023.




