---
title: "PSY6422 Data Management and Visualisation"
output:
  html_document:
    df_print: paged
    fig_width: 7
    fig_height: 6.37
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## The Growth of Woodland Area within the United Kingdom (1998-2023)

### Data Source

This project is focussed on visualising the increase in woodland within the United Kingdom, per country, from 1998 to 2023. The raw data for this project was retrieved from ['Forest Research'](https://www.forestresearch.gov.uk/tools-and-resources/statistics/data-downloads/), a branch of the Forestry Commission (FC) which collects data by totaling extractions made from multiple forest services operating within the United Kingdom. It principally gathers information on tree-related statistics, including but not limited to gathering longitudinal provisional data on the estimated amount of woodland area in the UK.

The “Woodland area, UK, 1998 to 2023” ODS file contained 5 sheets of actual data, though was also supplemented with cover, content and notes pages. In tandem, the pages containing data were annotated with further notes clarifying the origin of the data and different commissions involved in gathering it. The pages containing data split woodland area based on whether it was coniferous or deciduous, private or public sector, and totalled estimated woodland per country and in the United Kingdom as a whole.

```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Loading of necessary packages:

library(tidyverse)
library(ggplot2)
library(readODS)
library(here)
library(cowplot)

# Extraction of raw data from the ODS file

sheets<-c("England","Wales","Scotland","Northern Ireland")
countries_list<-list()
for (i in seq_along(sheets)){
  countries<-sheets[i]
  pathway<-paste0(here("raw_data", "area-timeseries-15jun23.ods"))
  countries_list[[countries]]<-read_ods(pathway, sheet = i+3)
} # this loop extracts all of the raw data and collates it into a data-frame

raw_extracted_data<-data.frame(countries_list)
rm(list=setdiff(ls(), "raw_extracted_data")) # removes now unnecessary variables to clean environment

# Glimpse of the raw data

head(raw_extracted_data)
```


### The Research Question

The main rationale for exploring this data was to elucidate whether there has been a growth in woodland area within the United Kingdom between 1998 and 2023. Furthermore, beyond just visualising the raw numeric increase, calculating the *percentage* increase per country would aid in contextualising whether countries are developing at similar rates, beyond just interpreting development based on the different amounts of reported woodland.

### Data Preparation

Based on this research question, the main variables of interest were the *Year* and the *total Woodland Area* (in thousand hectares) of each country, per year. Before continuing, lines of descriptive text needed to be removed in order to make the data more amenable to analysis.

```{r, echo=TRUE}
# Processing and cleaning of data

processed_extracted_data<-raw_extracted_data[-c(1:3),] # removes unnecessary text
names(processed_extracted_data)<-as.matrix(processed_extracted_data[1,]) # labels each column by their original titles
processed_extracted_data<-processed_extracted_data[-1,] # removes text names from the data

# Glimpse of the processed data

head(processed_extracted_data)
```

From the now processed data-frame, all of the data on the amount of woodland area needed to be tabulated to give a glimpse of how woodland has developed - as well as facilitate the calculation of percentage increase in total woodland from 1998 to 2023 for each country.

```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Extracts the Years and Amount of Woodland Area for each country

year_ending_March_31st<-as.numeric(processed_extracted_data$`Year ending 31 March`) # The Year
woodland_area<-select(processed_extracted_data, ends_with(" total (thousand ha)")) # Amount of Woodland Area
country_names<-c("England", "Wales", "Scotland", "Northern Ireland") # Country Names

for (i in country_names){
  assign(i, as.numeric(woodland_area[[paste0(i, " total (thousand ha)")]])) 
} # converts data to numeric for visualisation

# Tabulates change in woodland area per country from 1998 to 2023, and checks for any missing data

woodland_area_by_country<-data.frame(England,Wales,Scotland,`Northern Ireland`)
woodland_area_by_country<-woodland_area_by_country%>%
  rename(Northern_Ireland=`Northern.Ireland`)
table(is.na(woodland_area_by_country)) # no missing data!
country_names<-c("England", "Wales", "Scotland", "Northern_Ireland")

# Calculates percentage increase in woodland area per country from 1998 to 2023

percentage_results<-data.frame(country=character(),percentage_increase=numeric(),stringsAsFactors=FALSE) # creates an empty data-frame
for(country in country_names){
  min_v<-min(woodland_area_by_country[[country]])
  max_v<-max(woodland_area_by_country[[country]])
  percentage_increase<-(((max_v-min_v)/min_v)*100)
  percentage_results<-rbind(percentage_results, data.frame(country=country,percentage_increase=percentage_increase))
} # calculates the percentage increase from 1998 to 2023 for each country
percentage_results[1:4,2]<-round(percentage_results[1:4,2],2) # rounds the data to 2 decimal places

# Shows the data to allow a visual check for the validity and credibility of the results

woodland_area_by_country
percentage_results
```

As is viewable above, there has been evident growth of woodland within all countries of the United Kingdom from 1998 to 2023 - in tandem, calculating the percentage increase is especially beneficial for showing that in spite of smaller countries, namely Northern Ireland, having the lowest baseline (81 thousand hectares) and final (118 thousand hectares) woodland estimates, their *proportional* increase (45.68%) is quite notable.

### Data Visualisation

Finally, a visualisation of the growth of woodland area within the countries of the United Kingdom from 1998 to 2023 can be created, which displays this development both in terms of its numeric and proportional increase.

```{r, message=FALSE, echo=TRUE}
# Removes all the unnecessary variables, and creates a final data-frame amenable to the upcoming visualisation

rm(list=setdiff(ls(),c("processed_extracted_data","woodland_area_by_country","country_names","year_ending_March_31st", "plot","percentage_results")))
woodland_growth_over_time<-data.frame(
  country=c(rep("England",26),rep("Wales",26),rep("Scotland",26),rep("Northern Ireland",26)),
  woodland=c(woodland_area_by_country$England,woodland_area_by_country$Wales,
             woodland_area_by_country$Scotland,woodland_area_by_country$Northern_Ireland),
  year=c(year_ending_March_31st)
)

# Creates the mapping for the visualisation, and the figure path for saving it

mapping<-aes(x=year,y=woodland,colour=country) # creates the mapping for the visualisation
fig_path<-here("figs") # creates the necessary path for saving the figure

# Visualisation of the growth of woodland area within the United Kingdom, from 1998 to 2023

FinalPlot<-woodland_growth_over_time %>%
  ggplot(mapping=mapping)+
  geom_smooth(method="gam")+
  labs(x="Year (commencing from March 31st)",
       y="Woodland area (in thousand hectares)",
       colour="Country",
       title="The Growth of Woodland Area within the United Kingdom",
       subtitle="As annotated with percentage increase from 1998-2023",
       caption="Data retrieved from: Forest Research, 2023")+
  annotate("label",x=2010,y=1210,label=paste(percentage_results[1,2],"%"),colour="#EE0000",size=3,fontface="bold")+
  annotate("label",x=2010,y=375,label=paste(percentage_results[2,2],"%"),colour="#00CD00",size=3,fontface="bold")+
  annotate("label",x=2010,y=1450,label=paste(percentage_results[3,2],"%"),colour="#0000CD",size=3,fontface="bold")+
  annotate("label",x=2010,y=175,label=paste(percentage_results[4,2],"%"),colour="#FFA500",size=3,fontface="bold")+
  scale_colour_manual(values=c(England="#EE0000",Wales="#00CD00",
                               Scotland="#0000CD",`Northern Ireland`="#FFA500"))+
  scale_x_continuous(breaks=seq(1998,2023,5))+
  scale_y_continuous(breaks=seq(0,1500,150))+
  theme(panel.border=element_rect(colour="#8B7355",fill=NA,linewidth=2),
        panel.grid.minor=element_line(colour="#CAFF70",linewidth=0.5),
        panel.grid.major=element_line(colour="#CAFF70",linewidth=0.7),
        panel.background=element_rect(fill="#FFFFF0"),
        axis.line=element_line(linewidth=2,colour="#8B7355"),
        plot.subtitle=element_text(face="italic"),
        legend.title=element_blank(),
        legend.box.background=element_rect(colour="#8B7355"),
        legend.box.margin=margin(1,1,1,1),
        legend.key=element_rect(colour="#8B7355"))

FinalPlot
```

### Saving the Plot

```{r, message=FALSE, echo=TRUE}
# Saves the plot

filename<-paste("The Growth of Woodland Area in the UK from 1998 to 2023.png",sep="")
ggsave(file.path(fig_path,filename),plot=FinalPlot,width=7,height=6.37)
```

### Conclusion

From this plot, there is visual clarity that the growth of woodland area within countries should be interpreted both in the context of actual hectares of woodland area developed, as well as the proportional increase relative to the original amount of woodland within each country. This is because reporting either of these measures outside of the context of the other could be misleading. For example, interpreting that England or Scotland have made superior efforts to increase Woodland area compared to Northern Ireland, in the context that they have gained considerably higher *amounts* of woodland area from 1998-2023, neglects how Northern Ireland has shown greater proportional growth through a higher *percentage* increase in native woodland. Regardless of these differences, the data and accompanying visual can simply be interpreted on the promising notion that all countries within the United Kingdom have shown a growth in woodland area between 1998 and 2023. Further research could use data from the original file to map differences in the development of public or private sector woodland, or even split based on being deciduous or coniferous - in tandem, usage of other available data from the Forest Research service, such as how the amount of sawmills and wood production has changed from 1998-2023, could be interesting for investigating whether the rate of increasing woodland is sustainable compared to the rate of increasing wood production by sawmills in the UK.
