---
title: "Increase in the amount of Woodland within Countries of the United Kingdom,
  from 1998-2023"
author: "Jacob Moorcroft"
date: "2024-02-29"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This project is focussed on visualising the increase in woodland within the **United Kingdom**, as categorised by each *country*, from *1998* to *2023*. The data was retrieved from ['Forest Research'](https://www.forestresearch.gov.uk/tools-and-resources/statistics/data-downloads/), a branch of the Forestry Commission (FC) which collects data by totaling extractions made by multiple forest services operating in the United Kingdom. It principally gathers information on tree-related statistics, including but not limited to gathering longitudinal provisional data on the estimated amount of woodland area in the UK.

The “Woodland area, UK, 1998 to 2023” ODS file contained 5 tables, consisting of data on private and public woodland areas, of which coniferous or deciduous, within England, Wales, Scotland and Northern Ireland. The final table amalgamated this data into national totals across the United Kingdom, as shown in the raw data below.
	
```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(readODS)
RawData<-"raw_data/area-timeseries-15jun23.ods"
UKTotal<-read_ods(RawData,sheet=8)
head(UKTotal)

```

*Figure 1.* The raw data describing woodland area in the United Kingdom


As is visible within the raw data, it has been supplemented with a large amount of flavour text and descriptions. In tandem, it has the addition of dividing woodland area by public and privately owned land, as well as whether the area is primarily constituted of deciduous (i.e. oaks, cedars) or coniferous trees (i.e pines). For the purpose of this visualisation, the data needs to be cleaned to only include the *Year*, ranging from 1998-2023, and the *Total Woodland Area* per Year needs to be extracted from each sheet, per country and nationally.

The results of this screen and clean are below ...


```{r, message=FALSE, echo=FALSE}
library(tidyverse)
library(ggplot2)
library(readODS)

# extracts all of the raw data from the ODS and collates it into one data.frame
EnglandData<-read_ods("raw_data/area-timeseries-15jun23.ods",sheet=4) 
WalesData<-read_ods("raw_data/area-timeseries-15jun23.ods",sheet=5) 
ScotlandData<-read_ods("raw_data/area-timeseries-15jun23.ods",sheet=6) 
NIData<-read_ods("raw_data/area-timeseries-15jun23.ods",sheet=7) 
UKData<-read_ods("raw_data/area-timeseries-15jun23.ods",sheet=8)
TotalExtractedData<-data.frame(EnglandData,WalesData,ScotlandData,NIData,UKData)
rm(list=setdiff(ls(),"TotalExtractedData"))

TotalExtractedData<-TotalExtractedData[-c(1:3),] 
names(TotalExtractedData)<-as.matrix(TotalExtractedData[1,])
TotalExtractedData<-TotalExtractedData[-1,]

# extracts the total woodland area per country/the UK and the range of years, transforming into numerical variables
Year<-as.numeric(TotalExtractedData$`Year ending 31 March`)
Amount_of_Woodland<-select(TotalExtractedData, ends_with("total (thousand ha)"))

England<-as.numeric(Amount_of_Woodland$`England total (thousand ha)`)
Wales<-as.numeric(Amount_of_Woodland$`Wales total (thousand ha)`)
Scotland<-as.numeric(Amount_of_Woodland$`Scotland total (thousand ha)`)
NI<-as.numeric(Amount_of_Woodland$`Northern Ireland total (thousand ha)`)
UK_Total<-as.numeric(Amount_of_Woodland$`UK total (thousand ha)`)

# moves all of the relevant data into a data.frame
Woodland_By_Year<-data.frame(Year,England,Wales,Scotland,NI,UK_Total)
rm(list=setdiff(ls(),"Woodland_By_Year"))
```

```{r, echo=FALSE}
head(Woodland_By_Year)
```

*Figure 2.* The Amount of Woodland Area (in thousand hectacres) in the United Kingdom, total and per country, from 1998-2023